---
title: "HIV and AIDS Diagnoses in New York City (2016-2021)"
author: "Fahim Ahamed, Anthony Sinopoli, Fatima Nasyr"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # hides code when knit
  message = FALSE,
  warning = FALSE
) 
```

## Summary

The project examines how HIV and AIDS diagnoses have changed in New York City from 2010
to 2021. Using data from NYC Open Data, it explores patterns across time, geography, and
demographic groups to identify disparities in diagnosis rates and disease progression. The
analysis focuses on variations among the five boroughs and across gender and racial or ethnic
groups.

By highlighting which populations or areas experience higher rates, the study aims to provide
insights that can guide public health strategies such as awareness programs, prevention efforts,
and improved access to care and testing.

The exploratory data analysis (EDA) focuses on identifying key trends, distributions, and
patterns in the data. Using descriptive statistics and visualizations, it highlights how HIV and
AIDS diagnoses have shifted across New York City over time and across populations. The goal
was to develop a clear, data-driven understanding of the overall landscape before moving into
more detailed analyses.

### About the Data

The dataset used in this project was obtained from NYC Open Data, which provides publicly
available health statistics compiled by the New York City Department of Health and Mental
Hygiene (DOHMH). It contains yearly records of HIV and AIDS diagnoses among New York City
residents from 2010 to 2021, excluding 2014 and 2015, for which data were not reported.

Each record represents a specific combination of year, borough, neighborhood, sex, and
race/ethnicity, along with multiple indicators describing both counts and rates. These include:

● Total number of HIV diagnoses and corresponding rates per 100,000 population

● Number and proportion of concurrent HIV/AIDS diagnoses (individuals diagnosed with
both at the same time)

● Total number of AIDS diagnoses and corresponding rates per 100,000 population

More details about the dataset’s features and definitions are available on the official NYC Open
Data page.

### Data Preparation

The raw dataset from NYC Open Data required several cleaning and transformation steps
before exploratory analysis. First, the data were imported and inspected for duplicate entries,
missing values, and inconsistent formatting across variables. All numeric columns originally
stored as text were converted to numeric types for accurate computation.

To make the data consistent and analysis-ready:

●Duplicated rows were identified and removed, keeping only unique combinations of
year, borough, neighborhood, sex, and race/ethnicity.

● Line breaks and special characters were removed from categorical fields such as
RACE/ETHNICITY and Neighborhood (U.H.F).

● Rows with missing or suppressed values (shown as “NA”) were removed to ensure the
accuracy of summaries and visualizations.

● Data were filtered to include relevant combinations of attributes, such as cases where
borough, neighborhood, sex, and race/ethnicity were reported as “All,” depending on
the analysis focus.

## Exploratory Data Analysis

At the start of the analysis, we asked ourselves a key question: “What do we want to learn
from the data?”

The goal of the EDA was to explore how HIV and AIDS diagnoses vary across New York City by
examining the data from three different angles: yearly trends, geographic patterns (borough and
neighborhood), and demographic factors (sex and race/ethnicity).

Although the dataset spans 2010-2021, most visual analyses focus on 2016-2021 due to
missing or incomplete data in earlier years. This ensures consistent comparisons across
boroughs and demographic groups.

We wanted to see how these three views of the data relate to each other and whether changes
over the years also appear across different areas or groups. Looking at the data this way helped
us notice both the overall trends and the differences among specific populations.

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(forcats)
library(patchwork)
library(broom)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
library(rpart)
library(rpart.plot)
library(randomForest)
library(caret)
library(flextable)
```

```{r}
df <- read_csv('data/cleaned_hiv_data.csv')
df_no_agg <- read_csv("data/no_aggs_data.csv")
```

```{r}
hiv_summary <- df %>%
  filter(
    Borough == "All",
    Neighborhood == "All",
    SEX == "All",
    `RACE/ETHNICITY` != "All"
  ) %>% 
  group_by(`RACE/ETHNICITY`, YEAR) %>%
  summarise(
    avg_hiv_rate  = mean(`HIV DIAGNOSES PER 100,000 POPULATION`,  na.rm = TRUE),
    avg_aids_rate = mean(`AIDS DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hiv_summary,
       aes(x = YEAR, y = avg_hiv_rate, color = `RACE/ETHNICITY`)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2016:2021) +
  labs(
    title = "HIV Diagnosis Rates per 100,000 Population by Race/Ethnicity (2016-2021)",
    x = "Year",
    y = "Average HIV Diagnosis Rate",
    color = "Race/Ethnicity"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(
      face = "bold",
      size = 11,
      color = "black"
    ),
    axis.title.x = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    axis.text    = element_text(color = "black"),
    legend.title = element_text(face = "bold", color = "black")
  )
```

Across 2016-2021, the lines show clear differences in HIV rates by race/ethnicity. Groups at the
top of the plot maintain higher burdens throughout the period, while lower lines indicate
consistently smaller rates. Year-to-year changes are visible as small rises/dips; the relative
ordering between groups remains the main pattern to note.

```{r}
# Map setup
nyc_boroughs_sf =
  counties(state = 'NY', year = 2020, class = 'sf') %>%
  filter(NAME %in% c('Bronx', 'Kings', 'New York', 'Queens', 'Richmond'))

nyc_boroughs_sf$Borough = recode(
  nyc_boroughs_sf$NAME,
  'Bronx' = 'Bronx',
  'Kings' = 'Brooklyn',
  'New York' = 'Manhattan',
  'Queens' = 'Queens',
  'Richmond' = 'Staten Island'
)
```

```{r fig.width=10, fig.height=8}
df_race =
  df %>%
  filter(
    Borough != 'All',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` %in% c('Asian/Pacific Islander','Black','Latino/Hispanic','Other/Unknown','White')
  ) %>%
  group_by(Borough, `RACE/ETHNICITY`) %>%
  summarise(avg_rate = mean(`HIV DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE), .groups = 'drop')

race_map = nyc_boroughs_sf %>% left_join(df_race, by = 'Borough')

ggplot(race_map) +
  geom_sf(aes(fill = avg_rate), color = 'white', size = 0.6) +
  geom_sf_text(aes(label = Borough), color = 'black', size = 3.5, fontface = 'bold') +
  scale_fill_distiller(palette = 'OrRd', direction = 1, name = 'Avg Rate') +
  facet_wrap(~ `RACE/ETHNICITY`, ncol = 3) +
  theme_minimal() +
  labs(title = 'Average HIV Diagnosis Rates (Per 100K) by Race Across NYC Boroughs', x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        strip.text = element_text(size = 12, face = 'bold'),
        plot.title = element_text(hjust = 0.5, vjust = 5, size = 16, face = 'bold'))
```

The map plot shows strong racial differences in HIV diagnosis rates across NYC. Black
residents have the highest rates in every borough, especially in the Bronx, Manhattan, and
Brooklyn. Latino/Hispanic residents show moderately elevated rates, also highest in the Bronx
and Manhattan. Asian/Pacific Islander and White groups consistently have the lowest rates, with
minimal variation across boroughs. Overall, the visualization highlights a clear disparity, with
Black and Latino/Hispanic populations experiencing a disproportionately higher HIV burden
across the city.

```{r}
hiv_clean_sex <- df %>%
  mutate(YEAR = suppressWarnings(as.integer(YEAR))) %>%
  filter(
    !is.na(YEAR),
    YEAR >= 2016 & YEAR <= 2021,
    Borough == "All",
    Neighborhood == "All",
    `RACE/ETHNICITY` == "All"
  ) %>%
  mutate(SEX = str_to_title(SEX)) %>%
  filter(SEX %in% c("Female", "Male")) %>%
  filter(
    !is.na(`HIV DIAGNOSES PER 100,000 POPULATION`) |
      !is.na(`AIDS DIAGNOSES PER 100,000 POPULATION`)
  )

sex_year_summary <- hiv_clean_sex %>%
  group_by(SEX, YEAR) %>%
  summarise(
    avg_hiv_rate  = mean(`HIV DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE),
    avg_aids_rate = mean(`AIDS DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sex_year_summary, aes(x = YEAR, y = avg_hiv_rate, color = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "HIV Diagnosis Rates per 100,000 by Sex (2016-2021)",
    x = "Year",
    y = "Average HIV Diagnosis Rate",
    color = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")
```

The sex-based lines reveal how HIV rates change year by year for Females and Males. The
higher line indicates which sex carries more burden in a given year; the gap shows the size of
difference. Parallel movement suggests similar trends; diverging lines indicate changing
differences over time.

```{r}
ggplot(sex_year_summary, aes(x = YEAR, y = avg_aids_rate, color = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "AIDS Diagnosis Rates per 100,000 by Sex (2016-2021)",
    x = "Year",
    y = "Average AIDS Diagnosis Rate",
    color = "Sex"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")
```

AIDS rates by sex follow their own paths across the period. Compare the distance between the
lines to assess the sex gap and whether it widens or narrows. Differences from the HIV trend
can suggest variation in progression from HIV to AIDS.

```{r}
boro_race_avg <- df %>%
  mutate(YEAR = suppressWarnings(as.integer(YEAR))) %>%
  filter(
    !is.na(YEAR),
    YEAR >= 2016 & YEAR <= 2021,
    Borough != "All",
    Neighborhood == "All",
    SEX == "All",
    `RACE/ETHNICITY` != "All"
  ) %>%
  group_by(Borough, `RACE/ETHNICITY`) %>%
  summarise(
    mean_hiv_rate  = mean(`HIV DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE),
    mean_aids_rate = mean(`AIDS DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE),
    .groups = "drop"
  )

top_race_by_borough_hiv <- boro_race_avg %>%
  group_by(Borough) %>%
  slice_max(order_by = mean_hiv_rate,
            n = 1,
            with_ties = FALSE) %>%
  ungroup() %>% arrange(Borough)

top_race_by_borough_hiv %>%
  mutate(`RACE/ETHNICITY` = fct_reorder(`RACE/ETHNICITY`, mean_hiv_rate)) %>%
  ggplot(aes(
    x = mean_hiv_rate,
    y = fct_reorder(Borough, mean_hiv_rate),
    fill = `RACE/ETHNICITY`
  )) +
  geom_col() +
  labs(title = "Highest Race/Ethnicity by Borough (HIV Avg 2016-2021)",
       x = "Average HIV Diagnosis Rate per 100,000",
       y = "Borough",
       fill = "Race/Ethnicity") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"))
```

From 2016 to 2021, Black residents had the highest average HIV diagnosis rates in Manhattan,
the Bronx, Staten Island, and Brooklyn, while Latino/Hispanic residents had the highest rate in
Queens. The chart highlights clear borough-level differences, showing that the highest HIV rates
are concentrated within specific demographic groups in each borough.

```{r fig.width=10}
df_gender = df %>% 
  filter(Neighborhood == 'All', `RACE/ETHNICITY` == 'All', SEX %in% c('Male','Female')) %>%
  group_by(Borough, SEX) %>%
  summarise(avg_rate = mean(`HIV DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE), .groups = 'drop')

gender_map = nyc_boroughs_sf %>% left_join(df_gender, by = 'Borough')

p_male =
  ggplot(subset(gender_map, SEX == 'Male')) +
  geom_sf(aes(fill = avg_rate), color = 'white', size = 0.6) +
  geom_sf_text(aes(label = Borough), color = 'black', size = 3, fontface = 'bold') +
  scale_fill_distiller(palette = 'Blues', direction = 1, name = 'Avg HIV Rate') +
  theme_minimal() +
  labs(title = 'Male Average HIV Rate (2016-2021)', x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5, face = 'bold', size = 12))

p_female =
  ggplot(subset(gender_map, SEX == 'Female')) +
  geom_sf(aes(fill = avg_rate), color = 'white', size = 0.6) +
  geom_sf_text(aes(label = Borough), color = 'black', size = 3.5, fontface = 'bold') +
  scale_fill_distiller(palette = 'PuRd', direction = 1, name = 'Avg HIV Rate') +
  theme_minimal() +
  labs(title = 'Female Average HIV Rate (2016-2021)', x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5, face = 'bold', size = 12))

p_male + p_female +
  plot_annotation(title = 'Average HIV Rates by Gender Across NYC Boroughs (2016-2021)',
  theme = theme(plot.title = element_text(hjust = 0.5, vjust = 5, size = 16, face = 'bold')))
```

Male HIV rates are consistently higher than female rates across all NYC boroughs. The Bronx
shows the highest rates for both men and women, followed by Manhattan and Brooklyn, indicating
these areas carry the greatest HIV burden. Queens and Staten Island have the lowest average
rates for both genders. This shows a clear gender disparity, with men experiencing substantially
higher HIV rates while geographic differences across boroughs remain similar for both groups.

```{r fig.width=12, fig.height=5}
dist_race <- df %>%
  mutate(YEAR = suppressWarnings(as.integer(YEAR)),
         `RACE/ETHNICITY` = as.character(`RACE/ETHNICITY`)) %>%
  filter(
    !is.na(YEAR),
    YEAR >= 2016 & YEAR <= 2021,
    Borough != "All",
    Neighborhood == "All",
    SEX == "All",
    `RACE/ETHNICITY` != "All",
    !is.na(`HIV DIAGNOSES PER 100,000 POPULATION`) |
      !is.na(`AIDS DIAGNOSES PER 100,000 POPULATION`)
  ) %>%
  mutate(
    `RACE/ETHNICITY` = forcats::fct_reorder(
      `RACE/ETHNICITY`,
      `HIV DIAGNOSES PER 100,000 POPULATION`,
      .fun = median,
      .desc = TRUE,
      na.rm = TRUE
    )
  )

p_dist_race_hiv <- ggplot(
  dist_race,
  aes(x = `RACE/ETHNICITY`, y = `HIV DIAGNOSES PER 100,000 POPULATION`, fill = `RACE/ETHNICITY`)
) +
  geom_boxplot(outlier.shape = NA,
               width = 0.7,
               alpha = 0.9) +
  geom_jitter(width = 0.15,
              alpha = 0.25,
              size = 1) +
  coord_flip() +
  labs(title = "Distribution of HIV Diagnosis Rates by Race/Ethnicity\n(Borough-Level, 2016-2021)", x = "Race/Ethnicity", y = "HIV Diagnosis Rate per 100,000") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 10,
      color = "black",
      hjust = 0.5
    ),
    axis.title.x = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    legend.position = "none"
  )

p_dist_race_aids <- ggplot(
  dist_race,
  aes(x = `RACE/ETHNICITY`, y = `AIDS DIAGNOSES PER 100,000 POPULATION`, fill = `RACE/ETHNICITY`)
) +
  geom_boxplot(outlier.shape = NA,
               width = 0.7,
               alpha = 0.9) +
  geom_jitter(width = 0.15,
              alpha = 0.25,
              size = 1) +
  coord_flip() +
  labs(title = "Distribution of AIDS Diagnosis Rates by Race/Ethnicity\n(Borough-Level, 2016-2021)", x = "Race/Ethnicity", y = "AIDS Diagnosis Rate per 100,000") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 10,
      color = "black",
      hjust = 0.5
    ),
    axis.title.x = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    legend.position = "none"
  )

p_dist_race_hiv | p_dist_race_aids
```

The boxplots reveal clear disparities across race/ethnicity groups. Groups with higher medians
and wider IQRs show both greater overall burden and more variability across boroughs.
Comparing HIV vs AIDS panels shows whether the same groups lead on both measures or if
progression patterns differ.

```{r fig.width=10, fig.height=5}
dist_sex <- df %>%
  mutate(
    YEAR = suppressWarnings(as.integer(YEAR)),
    SEX = as.character(SEX)
  ) %>%
  filter(
    !is.na(YEAR), YEAR >= 2016 & YEAR <= 2021,
    Borough != "All", Neighborhood == "All",
    SEX != "All", `RACE/ETHNICITY` == "All",
    !is.na(`HIV DIAGNOSES PER 100,000 POPULATION`) | !is.na(`AIDS DIAGNOSES PER 100,000 POPULATION`)
  ) %>%
  mutate(
    SEX = forcats::fct_reorder(
      SEX, `HIV DIAGNOSES PER 100,000 POPULATION`, .fun = median, .desc = TRUE, na.rm = TRUE
    )
  )

p_dist_sex_hiv <- ggplot(
  dist_sex,
  aes(x = SEX, y = `HIV DIAGNOSES PER 100,000 POPULATION`, fill = SEX)
) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.9) +
  geom_jitter(width = 0.12, alpha = 0.25, size = 1) +
  labs(
    title = "Distribution of HIV Diagnosis Rates by Sex\n(Borough-Level, 2016-2021)",
    x = "Sex",
    y = "HIV Diagnosis Rate per 100,000"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "black", hjust = 0.5),
    axis.title.x = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    legend.position = "none"
  )

p_dist_sex_aids <- ggplot(
  dist_sex,
  aes(x = SEX, y = `AIDS DIAGNOSES PER 100,000 POPULATION`, fill = SEX)
) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.9) +
  geom_jitter(width = 0.12, alpha = 0.25, size = 1) +
  labs(
    title = "Distribution of AIDS Diagnosis Rates by Sex\n(Borough-Level, 2016-2021)",
    x = "Sex",
    y = "AIDS Diagnosis Rate per 100,000"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "black", hjust = 0.5),
    axis.title.x = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    axis.text.x  = element_text(color = "black"),
    axis.text.y  = element_text(color = "black"),
    legend.position = "none"
  )

p_dist_sex_hiv | p_dist_sex_aids
```

Distributions by sex show median differences and spread across boroughs. Wider boxes/longer
whiskers signal greater variability across borough-year observations. Side-by-side HIV vs AIDS
panels help assess whether the sex gap is consistent at diagnosis and at AIDS stage.

```{r fig.width=9}
df_borough <- df %>%
  filter(Borough %in% c('Bronx', 'Brooklyn', 'Manhattan', 'Queens', 'Staten Island'),
         Neighborhood == 'All',
         SEX == 'All',
         `RACE/ETHNICITY` == 'All') %>%
  select(Borough, YEAR,
         `HIV DIAGNOSES PER 100,000 POPULATION`,
         `AIDS DIAGNOSES PER 100,000 POPULATION`) %>%
  pivot_longer(
    cols = c(`HIV DIAGNOSES PER 100,000 POPULATION`,
             `AIDS DIAGNOSES PER 100,000 POPULATION`),
    names_to = 'Measure',
    values_to = 'Rate'
  )

ggplot(df_borough, aes(x = YEAR, y = Rate, color = Measure)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c('HIV DIAGNOSES PER 100,000 POPULATION' = 'firebrick',
                                'AIDS DIAGNOSES PER 100,000 POPULATION' = 'steelblue'),
                     labels = c('AIDS Rate', 'HIV Rate')) +
  facet_wrap(~Borough, ncol = 3) +
  labs(
    title = 'HIV and AIDS Diagnosis Rates by Borough (2016-2021)',
    x = 'Year',
    y = 'Rate per 100,000 Population',
    color = ''
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = 'bold', hjust = 0.5),
    legend.position = 'top',
    strip.text = element_text(face = 'bold')
  )
```

Overall, a general decline in HIV diagnoses until 2020, where a spike can be seen from 2020-2021
(except in Brooklyn). The Bronx is the borough with the highest diagnoses, and Staten Island with the
lowest. Again, the Bronx has the highest AIDS diagnoses with a spike from 2020-2021. Manhattan
has been relatively stable, and Brooklyn has even seen a slight decline. Queens and Staten Island
have the lowest diagnoses, but have seen an increase from 2020-2021.

```{r}
neighborhood_summary <- df %>% 
  filter(
    Borough != "All",
    Neighborhood != "All",
    SEX == "All",
    `RACE/ETHNICITY` == "All"
  ) %>% 
  group_by(YEAR, Borough, Neighborhood)

ggplot(data = neighborhood_summary, aes(`HIV DIAGNOSES PER 100,000 POPULATION`, `AIDS DIAGNOSES PER 100,000 POPULATION`)) +
  geom_point(aes(color = Borough, shape = Borough),
             alpha = 0.8,
             size = 2) +
  facet_wrap(~ Borough) +
  labs(
    title = 'HIV and AIDS Diagnoses by Borough',
    x = 'HIV Diagnoses (per 100,000)',
    y = 'AIDS Diagnoses (per 100,000)'
  ) +
  scale_color_brewer(palette = 'Dark2') +
  theme_minimal()
```

The plot shows a clear positive relationship between HIV and AIDS diagnosis rates across all five
boroughs, meaning that areas with higher HIV rates also tend to have higher AIDS rates. The Bronx
stands out with the highest overall rates, while Queens and Staten Island have noticeably lower
values. This suggests that the disease burden is not evenly distributed across the city.

```{r}
concurrent_borough <- df %>% 
  filter(
    Borough != "All",
    Neighborhood == "All",
    SEX == "All",
    `RACE/ETHNICITY` == "All"
  ) %>% 
  group_by(YEAR, Borough)

ggplot(data = concurrent_borough, aes(YEAR,
                                      `PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`,
                                      color = Borough)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Proportion of Concurrent HIV/AIDS Diagnoses by Borough", 
    x = "",
    y = "Proportion of Diagnoses"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

The proportion of concurrent HIV/Aids diagnoses is a measure of late diagnoses of HIV, where the
patient is diagnosed for HIV, then AIDS at the same time or shortly after. Manhattan has seen an
increase in late diagnoses.

```{r fig.width=8}
bronx_all <- df %>% 
  filter(
    Borough == "Bronx",
    Neighborhood == "All",
    SEX != "All",
    `RACE/ETHNICITY` != "All"
  ) %>% 
  group_by(YEAR)

ggplot(data = bronx_all, aes(x = YEAR,
                             y = `HIV DIAGNOSES PER 100,000 POPULATION`,
                             group = SEX,
                             shape = SEX,
                             color = SEX)) +
  geom_line(size = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ `RACE/ETHNICITY`) +
  labs(
    title = 'Bronx HIV Diagnoses by Race/Ethnicity and Sex',
    x = '',
    y = 'HIV Diagnoses (per 100,000)'
  ) +
  scale_shape_manual(
    values = c(
      'Female' = 16,  # circle
      'Male' = 15     # square
    )
  ) +
  scale_color_manual(
    values = c(
      'Female' = '#e75480',  # pink/red tone
      'Male' = '#0072B2'     # blue
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.85, 0.20),
    legend.background = element_rect(fill = alpha('white', 0.8), color = NA),
    legend.box.background = element_blank()
  )
```

Taking a more in-depth look at the Bronx, males have a higher rate of HIV diagnoses than females,
when there is data (missing some female data). Black males have the highest overall HIV diagnoses,
with Latino/Hispanic males having the second most. Males in the other races have similar diagnoses
to one another.

Note: Some female data points (Asian and Other) are missing because their reported values were
either zero or marked as NA in the dataset. The Other/Unknown race female has one data point,
which is for 2019.

```{r fig.width=8}
df_long <- df %>%
  filter(Neighborhood != 'All',
         `RACE/ETHNICITY` == 'All',
         SEX == 'All') %>%
  select(Borough,
         `HIV DIAGNOSES PER 100,000 POPULATION`,
         `AIDS DIAGNOSES PER 100,000 POPULATION`) %>%
  pivot_longer(
    cols = c(`HIV DIAGNOSES PER 100,000 POPULATION`,
             `AIDS DIAGNOSES PER 100,000 POPULATION`),
    names_to = 'Type',
    values_to = 'Rate'
  )

df_long$Type <- recode(df_long$Type,
                       'HIV DIAGNOSES PER 100,000 POPULATION' = 'HIV Diagnoses',
                       'AIDS DIAGNOSES PER 100,000 POPULATION' = 'AIDS Diagnoses')

ggplot(df_long, aes(x = Borough, y = Rate, fill = Borough)) +
  geom_boxplot() +
  facet_wrap(~ Type, ncol = 2) +
  labs(x = 'Borough', y = 'Rate per 100,000',
       title = 'Distribution of HIV and AIDS Diagnosis Rates by Borough') +
  theme_minimal() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5),
        legend.position = 'none')
```

This boxplot compares the distribution of HIV and AIDS diagnosis rates across New York City’s five
boroughs. In both panels, the Bronx stands out with the highest median rates and widest spread,
indicating higher and more variable neighborhood-level burden. Brooklyn and Manhattan follow with
moderately high and overlapping ranges, while Queens and Staten Island consistently show lower
rates and narrower distributions.

The pattern suggests that HIV and AIDS diagnoses are not evenly distributed across the city. The
Bronx remains the most affected borough, while outer boroughs like Queens and Staten Island
experience relatively lower and more stable rates.

```{r}
ggplot(df[df$Neighborhood != 'All' &
          df$`RACE/ETHNICITY` == 'All' &
          df$SEX == 'All', ],
       aes(x = `HIV DIAGNOSES PER 100,000 POPULATION`)) +
  geom_histogram(bins = 30, fill = 'steelblue', color = 'white') +
  geom_vline(aes(xintercept = mean(`HIV DIAGNOSES PER 100,000 POPULATION`, na.rm = TRUE)),
             color = 'red', linewidth = 1.1, linetype = 'dashed') +
  labs(x = 'HIV Diagnoses per 100,000', y = 'Count',
       title = 'Distribution of HIV Rates Across NYC Neighborhoods') +
  theme_minimal() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))
```

This histogram shows how HIV diagnosis rates are distributed across New York City neighborhoods.
Most neighborhoods have relatively low rates, clustered below the citywide average. The distribution
is right-skewed, meaning only a few neighborhoods experience much higher rates, which raises the
overall mean.

This pattern further confirms that the HIV burden in NYC is unevenly distributed, concentrated in a
limited number of high-incidence neighborhoods while the majority of areas maintain lower rates. It
highlights persistent geographic disparities, where targeted prevention and testing efforts may be
most needed in those higher-rate communities.

```{r fig.height=6, fig.width=10}
df_years <- df %>%
  filter(
    Borough == 'All' &
    Neighborhood == 'All' &
    SEX == 'All' &
    `RACE/ETHNICITY` == 'All'
  )

df_burden <- df_years %>%
  mutate(
    YEAR = as.factor(YEAR),
    Total_Diagnoses = `TOTAL NUMBER OF HIV DIAGNOSES` + `TOTAL NUMBER OF AIDS DIAGNOSES`
  )

ggplot(df_burden, aes(x = YEAR, y = Total_Diagnoses)) +
  geom_col(fill = 'steelblue', alpha = 0.8) +
  geom_text(aes(label = Total_Diagnoses), vjust = -0.8, size = 3.5) +
  labs(
    title = 'Total HIV and AIDS Diagnoses in NYC (2016-2021)',
    x = 'Year',
    y = 'Total New Diagnoses'
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = 'bold', hjust = 0.5)
  )
```

This chart shows the total burden of HIV and AIDS in New York City from 2016 to 2021, measured by
the number of new diagnoses each year. The overall downward trend indicates that fewer people are
entering the HIV care system, reflecting progress in prevention, early detection, and long-term
treatment efforts. The dip in 2020 suggests a temporary decline in testing or reporting during the
pandemic, while the slight increase in 2021 points to a return to normal healthcare activity. Overall,
the data highlights a steady reduction in the city’s HIV/AIDS burden over time.

```{r}
df_change <- df_years %>%
  arrange(YEAR) %>%
  mutate(
    HIV_change = (`HIV DIAGNOSES PER 100,000 POPULATION` - lag(`HIV DIAGNOSES PER 100,000 POPULATION`)) /
                 lag(`HIV DIAGNOSES PER 100,000 POPULATION`) * 100,
    AIDS_change = (`AIDS DIAGNOSES PER 100,000 POPULATION` - lag(`AIDS DIAGNOSES PER 100,000 POPULATION`)) /
                  lag(`AIDS DIAGNOSES PER 100,000 POPULATION`) * 100
  )

df_change <- df_change %>% filter(YEAR > 2016)

ggplot(df_change, aes(x = YEAR)) +
  geom_line(aes(y = HIV_change, color = 'HIV Rate Change'), size = 1.2) +
  geom_point(aes(y = HIV_change, color = 'HIV Rate Change'), size = 2) +
  geom_line(aes(y = AIDS_change, color = 'AIDS Rate Change'), size = 1.2) +
  geom_point(aes(y = AIDS_change, color = 'AIDS Rate Change'), size = 2) +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'gray50') +
  scale_color_manual(values = c('HIV Rate Change' = 'steelblue', 'AIDS Rate Change' = 'firebrick')) +
  labs(
    title = 'Year-over-Year Change in HIV and AIDS Rates (2017-2021)',
    x = 'Year',
    y = 'Percent Change from Previous Year (%)',
    color = ''
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = 'bold', hjust = 0.5),
    legend.position = 'top'
  )
```

This chart shows the year-over-year percentage change in HIV and AIDS diagnosis rates in New York
City from 2017 to 2021. Both rates declined consistently through 2019, reflecting steady progress in
reducing new infections. A sharp drop in 2020 suggests significant disruptions in testing and care
access during the pandemic. The strong rebound in 2021 likely represents resumed testing and
diagnosis rather than a true increase in new cases. Overall, the pattern shows long-term improvement
with short-term fluctuations linked to external healthcare disruptions.

```{r fig.width=12, fig.height=6}
df_faceted <- df_years %>%
  select(YEAR,
         `TOTAL NUMBER OF CONCURRENT HIV/AIDS DIAGNOSES`,
         `PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`) %>%
  pivot_longer(
    cols = c(`TOTAL NUMBER OF CONCURRENT HIV/AIDS DIAGNOSES`,
             `PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`),
    names_to = 'Measure',
    values_to = 'Value'
  )

df_faceted$Measure[df_faceted$Measure == 'TOTAL NUMBER OF CONCURRENT HIV/AIDS DIAGNOSES'] <- 'Total Concurrent HIV/AIDS Diagnoses'
df_faceted$Measure[df_faceted$Measure == 'PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES'] <- 'Proportion of Concurrent HIV/AIDS Diagnoses (%)'

ggplot(df_faceted, aes(x = YEAR, y = Value)) +
  geom_line(color = 'firebrick', size = 1.2) +
  geom_point(color = 'firebrick', size = 2) +
  geom_text(aes(label = round(Value, 1)), vjust = -0.8, size = 3.5) +
  facet_wrap(~Measure, scales = 'free_y', ncol = 2) +
  labs(
    title = 'Concurrent HIV/AIDS Diagnoses in NYC (2016-2021)',
    x = 'Year',
    y = ''
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = 'bold', hjust = 0.5),
    strip.text = element_text(face = 'bold')
  )
```

This figure shows how often people in New York City were diagnosed with HIV and AIDS at the same
time between 2016 and 2021. The left chart shows the share of new HIV cases that were already
AIDS at diagnosis, while the right chart shows the total number of such cases each year. Together,
these trends suggest that while the number of concurrent cases dropped, the share of HIV diagnoses
that were already AIDS at detection grew slightly. This means fewer people were diagnosed overall,
but a larger portion of them had advanced infection, indicating slower or delayed testing in recent
years. The small rise in both metrics in 2021 likely reflects the rebound of HIV testing and diagnosis
after the pandemic’s disruptions.

```{r fig.height=6, fig.width=10}
df_gap <- df_years %>%
  mutate(Gap = `HIV DIAGNOSES PER 100,000 POPULATION` - `AIDS DIAGNOSES PER 100,000 POPULATION`)

ggplot(df_gap, aes(x = YEAR, y = Gap)) +
  geom_line(color = 'steelblue', size = 1.2) +
  geom_point(color = 'steelblue', size = 2) +
  geom_text(aes(label = round(Gap, 1)), vjust = -0.8, size = 3.5) +
  labs(
    title = 'Gap Between HIV and AIDS Diagnosis Rates in NYC (2016-2021)',
    x = 'Year',
    y = 'Rate Difference (per 100,000 people)'
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = 'bold', hjust = 0.5)
  )
```

This chart shows the difference between HIV and AIDS diagnosis rates in New York City from 2016 to
2021. The gap between the two rates narrows over time. The shrinking gap aligns with the earlier
finding that a larger share of new HIV diagnoses were already at an advanced stage, suggesting the
two measures are moving closer together over time.

```{r fig.width=10}
df_most_affected <- do.call(rbind, by(df, df$YEAR, function(x) {
  x <- x[x$Neighborhood != 'All' &
         x$`RACE/ETHNICITY` == 'All' &
         x$SEX == 'All' &
         x$Borough != 'All', ]
  hiv_top <- x[which.max(x$`HIV DIAGNOSES PER 100,000 POPULATION`), ]
  aids_top <- x[which.max(x$`AIDS DIAGNOSES PER 100,000 POPULATION`), ]
  hiv_top$Type <- 'HIV Diagnoses'
  aids_top$Type <- 'AIDS Diagnoses'
  rbind(hiv_top, aids_top)
}))

df_most_affected$Type <- factor(df_most_affected$Type,
                                levels = c('HIV Diagnoses', 'AIDS Diagnoses'))

ggplot(df_most_affected,
       aes(x = factor(YEAR),
           y = ifelse(Type == 'HIV Diagnoses',
                      `HIV DIAGNOSES PER 100,000 POPULATION`,
                      `AIDS DIAGNOSES PER 100,000 POPULATION`),
           fill = Borough)) +
  geom_col() +
  geom_text(aes(y = ifelse(Type == 'HIV Diagnoses',
                           `HIV DIAGNOSES PER 100,000 POPULATION` / 2,
                           `AIDS DIAGNOSES PER 100,000 POPULATION` / 2),
                label = Neighborhood),
            angle = 90, vjust = 0.5, hjust = 0.5, size = 4, color = 'black') +
  facet_wrap(~ Type, ncol = 2, scales = 'free_y') +
  labs(x = 'Year', y = 'Rate per 100,000',
       fill = 'Borough',
       title = 'Most Affected Neighborhood Each Year by HIV and AIDS Diagnosis Rates') +
  theme_minimal() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))
```

Across the years, the Bronx dominates as the borough with the most affected neighborhoods for both
HIV and AIDS diagnoses. High Bridge-Morrisania and Crotona-Tremont appear repeatedly, showing
that certain Bronx neighborhoods continue to carry the highest burden. Brooklyn’s East New York is
the only neighborhood outside the Bronx that stands out, appearing once in 2020 for HIV diagnoses.
This pattern suggests that while rates vary slightly by year, the Bronx consistently faces the highest
concentration of new and advanced HIV cases citywide.

### Summary of Key Findings

Overall, the analysis shows that HIV and AIDS diagnoses in New York City have declined over time,
reflecting progress in prevention and care, though disparities remain across groups and locations. The
Bronx continues to show the highest rates, while Queens and Staten Island report the lowest. Black
and Latino/Hispanic residents experience the greatest overall burden, and males consistently have
higher diagnosis rates than females. The temporary dip in 2020 and rebound in 2021 likely reflect
testing disruptions during the pandemic. Together, these findings highlight steady citywide
improvement but persistent demographic and geographic gaps that remain important for public health
planning.

## Hypothesis Testing

### 1. Was the 2020 HIV Diagnosis Total Significantly Lower Than Expected?

To assess whether the drop in HIV diagnoses observed in 2020 was unusually large compared to pre-
pandemic levels, a one-sample t-test was performed. The goal is to compare the 2020 total number of
HIV diagnoses to the average annual total from the four pre-COVID years (2016 to 2019).

#### Hypotheses

Let μ represent the true average annual number of HIV diagnoses in the pre-COVID period.

• H₀: μ = HIV_2020
The 2020 total is consistent with the pre-COVID average.

• H₁: μ > HIV_2020
The pre-COVID average is higher than the 2020 value, indicating a meaningful decline in
2020.

#### Assumptions

• The yearly totals from 2016 to 2019 are independent.

• The sample size is small (n = 4), so the t-distribution is appropriate.

• No extreme outliers are present among the pre-COVID yearly totals.

```{r}
df_years <- df %>%
  filter(
    Borough == 'All',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` == 'All'
  ) %>%
  select(-Borough, -Neighborhood, -SEX, -`RACE/ETHNICITY`)

pre_2020 <- df_years$`TOTAL NUMBER OF HIV DIAGNOSES`[df_years$YEAR %in% 2016:2019]

rate_2020 <- df_years$`TOTAL NUMBER OF HIV DIAGNOSES`[df_years$YEAR == 2020]

t.test(pre_2020, mu = rate_2020, alternative = 'greater')
```

The one-sample t-test indicates that HIV diagnoses in 2020 were much lower than what we typically
saw before the pandemic. The average number of cases from 2016-2019 was clearly higher than the
2020 value of 1,394, and the p-value (0.005) shows that this drop is unlikely to be due to normal year-
to-year variation.

Based on this result, we conclude that the decline in 2020 was statistically significant and likely
reflects the impact of COVID-19 on HIV testing and reporting.

```{r}
ggplot(df_years, aes(x = YEAR, y = `TOTAL NUMBER OF HIV DIAGNOSES`)) +
  
  geom_point(size = 3, color = 'black') +
  geom_line(linewidth = 1.1, color = 'steelblue') +
  
  geom_point(
    data = subset(df_years, YEAR == 2020),
    aes(x = YEAR, y = `TOTAL NUMBER OF HIV DIAGNOSES`),
    color = 'red', size = 4
  ) +
  
  scale_y_continuous(
    limits = c(1000, 2500),
    breaks = seq(1000, 2500, 250),
    expand = c(0, 0)
  ) +
  
  labs(
    title = 'Yearly HIV Diagnoses in NYC (2016-2021)',
    x = 'Year',
    y = 'Total HIV Diagnoses'
  ) +
  theme_minimal()
```

### 2. Are Bronx HIV Rates Consistently Higher Than the NYC Average?

To examine whether the Bronx consistently reports higher HIV diagnosis rates than the overall New
York City average, a paired t-test was conducted. For each year from 2016 to 2021, the Bronx HIV
rate was matched with the corresponding citywide rate, creating yearly paired observations.

#### Hypotheses

Let 𝑑 represent the difference between the Bronx HIV rate and the NYC-wide rate for each year.

• H₀: mean(d) = 0
On average, Bronx HIV rates are equal to the citywide rates.

• H₁: mean(d) > 0
Bronx HIV rates are higher on average, indicating a consistent elevation relative to the NYC
average.

#### Assumptions

• Each year’s Bronx-NYC rate difference is independent.

• The distribution of the differences is approximately normal.

• Each year forms a valid matched pair for comparison.

```{r, include=FALSE}
bronx <- df %>%
  filter(
    Borough == 'Bronx',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` == 'All'
  )

nyc <- df %>%
  filter(
    Borough == 'All',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` == 'All'
  )

bronx_rates <- bronx$`HIV DIAGNOSES PER 100,000 POPULATION`
nyc_rates <- nyc$`HIV DIAGNOSES PER 100,000 POPULATION`
diff_rates <- bronx_rates - nyc_rates

data.frame(
  YEAR = bronx$YEAR,
  Bronx_Rate = bronx_rates,
  NYC_Rate = nyc_rates,
  Difference = diff_rates
)
```

```{r}
t.test(bronx_rates, nyc_rates,
       paired = TRUE,
       alternative = 'greater')
```


The paired t-test shows that HIV diagnosis rates in the Bronx were consistently higher than the NYC-
wide rate from 2016 to 2021. On average, the Bronx rate was about 9.7 cases per 100,000 above the
citywide level. The test result (t = 7.81, p = 0.00028) and the one-sided 95% confidence interval (7.22,
∞) both indicate a clear and statistically significant difference. Overall, the evidence strongly supports
that the Bronx remains a higher-burden area compared to the rest of New York City.

```{r}
ggplot() +
  geom_line(data = bronx, aes(x = YEAR, y = `HIV DIAGNOSES PER 100,000 POPULATION`, 
                              color = 'Bronx'), linewidth = 1.2) +
  geom_point(data = bronx, aes(x = YEAR, y = `HIV DIAGNOSES PER 100,000 POPULATION`, 
                               color = 'Bronx'), size = 3) +
  
  geom_line(data = nyc, aes(x = YEAR, y = `HIV DIAGNOSES PER 100,000 POPULATION`, 
                            color = 'NYC Average'), linewidth = 1.2) +
  geom_point(data = nyc, aes(x = YEAR, y = `HIV DIAGNOSES PER 100,000 POPULATION`, 
                             color = 'NYC Average'), size = 3) +
  
  scale_color_manual(values = c('Bronx' = 'red', 'NYC Average' = 'blue')) +
  
  labs(
    title = 'HIV Diagnosis Rates: Bronx vs NYC Average (2016-2021)',
    x = 'Year',
    y = 'HIV Diagnoses per 100,000 Population',
    color = ''
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 1)
  )
```

### 3. Does Queens have a higher proportion of concurrent HIV/AIDS diagnoses than the high-burden boroughs, even though it has fewer total HIV cases?

To examine whether Queens experiences higher late-stage HIV diagnoses despite having fewer total
HIV cases, a two-sample t-test was conducted. Each year from 2016 to 2021, the concurrent
diagnosis proportion for Queens was compared to the combined concurrent proportions of the high-
burden boroughs (Bronx, Brooklyn, and Manhattan). This produced six yearly observations for
Queens and eighteen for the high-burden group.

#### Hypotheses

Let 𝜇𝑄 represent the mean concurrent HIV/AIDS diagnosis proportion in Queens, and 𝜇𝐻 represent
the mean concurrent diagnosis proportion in the high-burden boroughs.

• H₀: μ_Q ≤ μ_H
On average, Queens has concurrent proportions that are the same or lower than the high-
burden boroughs.

• H₁: μ_Q > μ_H
Queens has a higher concurrent proportion, indicating more frequent late-stage diagnoses
despite being a low-burden borough.

#### Assumptions

• Borough-year observations are independent.

• The yearly concurrent proportions for Queens (n = 6) and the high-burden boroughs (n = 18)
are approximately normally distributed.

• Unequal sample sizes and variances are acceptable under Welch’s t-test.

```{r}
df_years <- df %>%
  filter(
    Borough != 'All',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` == 'All'
  ) %>%
  select(-Neighborhood, -SEX, -`RACE/ETHNICITY`)

high_burden <- c('Bronx', 'Brooklyn', 'Manhattan')

high_vals <- df_years$`PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`[
  df_years$Borough %in% high_burden
]

queens_vals <- df_years$`PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`[
  df_years$Borough == 'Queens'
]
  
t.test(queens_vals, high_vals, alternative = 'greater')
```

The one-sided t-test shows strong evidence that Queens has a higher proportion of concurrent
HIV/AIDS diagnoses than the high-burden boroughs (p = 0.0074). Since the p-value is well below
0.05, we reject the null hypothesis and conclude that Queens experiences significantly more late-
stage diagnoses. This suggests that, despite having fewer overall HIV cases, Queens faces greater
challenges with early detection and timely testing compared to the high-burden boroughs.

```{r}
df_years <- df_years %>%
  mutate(
    group_label = case_when(
      Borough == 'Queens' ~ 'Queens (Low-burden)',
      Borough == 'Staten Island' ~ 'Staten Island (Low-burden)',
      TRUE ~ 'High-burden Boroughs'
    )
  )

ggplot(df_years, aes(
  x = YEAR,
  y = `PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES AMONG ALL HIV DIAGNOSES`,
  group = Borough,
  color = group_label,
  alpha = group_label
)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +

  # Custom colors
  scale_color_manual(
    values = c(
      'Queens (Low-burden)' = 'red',
      'Staten Island (Low-burden)' = '#3CB371',   # green-ish
      'High-burden Boroughs' = 'grey70'
    )
  ) + 
  # Custom alpha levels
  scale_alpha_manual(
    values = c(
      'Queens (Low-burden)' = 1,
      'Staten Island (Low-burden)' = 0.4,       
      'High-burden Boroughs' = 0.9
    )
  ) +

  labs(
    title = 'Concurrent HIV/AIDS Diagnosis Proportion by Borough (2016-2021)',
    x = 'Year',
    y = 'Concurrent Diagnosis Proportion',
    color = 'Borough Category',
    alpha = 'Borough Category'
  ) +

  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.35)
  )
```

## Modeling

Yearly Trends: We fit a linear model to examine how total HIV diagnoses changed from 2016 to 2021
and to quantify whether the downward trend seen in the data is statistically meaningful.

```{r}
df_years <- df %>%
  filter(
    Borough == 'All',
    Neighborhood == 'All',
    SEX == 'All',
    `RACE/ETHNICITY` == 'All'
  ) %>%
  select(-Borough, -Neighborhood, -SEX, -`RACE/ETHNICITY`)

model1 <- lm(`TOTAL NUMBER OF HIV DIAGNOSES` ~ YEAR, data = df_years)

summary(model1)
```

The linear model shows a clear and statistically significant downward trend in total HIV diagnoses
from 2016 to 2021. The estimated slope is −154.06, indicating that diagnoses decreased by roughly
154 cases per year on average. This decline is statistically significant (p ≈ 0.007), meaning it is
unlikely to be due to chance. The model explains a large portion of the year-to-year variation (R² =
0.866), suggesting that the overall decline is strong and consistent across the observed period.

```{r}
df_years$pred <- predict(model1)

ggplot(df_years, aes(YEAR, `TOTAL NUMBER OF HIV DIAGNOSES`)) +
  geom_point(size = 3) +
  geom_line(aes(y = pred), color = "red", linewidth = 1.2) +
  theme_minimal()
```

To check whether the trend in HIV diagnoses followed a curved pattern rather than a straight-line
decline, we fit a quadratic model that includes a YEAR² term. This helps us test whether a nonlinear
pattern provides a better explanation of the data.

```{r}
model_poly <- lm(`TOTAL NUMBER OF HIV DIAGNOSES` ~ YEAR + I(YEAR^2), data = df_years)
summary(model_poly)
```

The quadratic model does not provide evidence of a curved trend. Neither the YEAR term nor the
YEAR² term is significant (p ≈ 0.46), meaning the data do not support any meaningful nonlinear
pattern. Even though the R² is high, the added quadratic term does not improve the model. Overall,
the trend remains essentially linear.

```{r}
df_years$pred_poly <- predict(model_poly)

ggplot(df_years, aes(YEAR, `TOTAL NUMBER OF HIV DIAGNOSES`)) +
  geom_point(size = 3) +
  geom_line(aes(y = pred_poly), color = "blue", linewidth = 1.2) +
  theme_minimal()
```

The fitted curve from the quadratic model looks nearly identical to a straight line, reflecting the fact
that the quadratic term is not significant. The predicted values follow the same steady downward
pattern seen in the data, with no noticeable curvature. This visual pattern reinforces the model result:
the trend in HIV diagnoses is effectively linear over this period.

```{r}
df_hiv_pred <- df_years %>% 
  select(YEAR, `TOTAL NUMBER OF HIV DIAGNOSES`, pred, pred_poly)

ggplot(df_hiv_pred, aes(x = YEAR)) +
  geom_point(aes(y = `TOTAL NUMBER OF HIV DIAGNOSES`, color = "Actual"),
             size = 3) +
  
  geom_line(aes(y = pred, color = "Linear Model"),
            linewidth = 1.2) +
  
  geom_line(aes(y = pred_poly, color = "Polynomial Model"),
            linewidth = 1.2) +
  
  scale_color_manual(
    values = c(
      "Actual" = "black",
      "Linear Model" = "red",
      "Polynomial Model" = "blue"
    ),
    name = "Legend"
  ) +
  
  labs(
    title = "Actual Data vs Linear and Polynomial Model Predictions",
    x = "Year",
    y = "Total Number of HIV Diagnoses"
  ) +
  
  theme_minimal()
```

The plot compares the actual HIV diagnoses with predictions from the linear and quadratic models.
When viewed side by side, the polynomial curve shows a slight curvature, but it does not meaningfully
change the overall trend. Both fitted lines closely follow the steady downward pattern in the data,
reinforcing that the decline is essentially linear.

### Testing Extrapolation for 2022-2024

To explore how well our model extends beyond the observed data, we checked the most recent NYC
HIV Surveillance Annual Reports for 2022 and 2023 and compared their published totals with our
model’s extrapolated predictions. The 2024 report is not yet available.

It is important to note that the annual reports do not perfectly match the values in our dataset. For
example, the reports list 1,594 cases for 2021 (our dataset: 1,592), 1,396 cases for 2020 (dataset:
1,394), and 1,772 cases for 2019 (dataset: 1,762). These small discrepancies likely reflect data
revisions or reporting differences between sources.

Nonetheless, we were curious to see whether the overall trend observed from 2016 to 2021 continued
in recent years, and whether our fitted model could reasonably approximate the direction of that trend.

```{r}
future_years <- data.frame(
  YEAR = c(2022, 2023),
  'TOTAL NUMBER OF HIV DIAGNOSES' = c(1624, 1686)
)

future_years$pred_linear <- predict(model1, future_years)
future_years$pred_poly   <- predict(model_poly, future_years)

names(future_years)[names(future_years) == 'TOTAL.NUMBER.OF.HIV.DIAGNOSES'] <- 
  'TOTAL NUMBER OF HIV DIAGNOSES'

names(df_hiv_pred)[names(df_hiv_pred) == 'pred'] <- 'pred_linear'

df_hiv_pred <- rbind(df_hiv_pred, future_years)

ggplot(df_hiv_pred, aes(x = YEAR)) +
  geom_point(aes(y = `TOTAL NUMBER OF HIV DIAGNOSES`, color = 'Actual'), size = 2) +
  
  geom_line(aes(y = pred_linear, color = 'Linear Model'), linewidth = 1.5, alpha=0.35) +
  geom_point(aes(y = pred_linear, color = 'Linear Model'), size = 2, alpha=0.5) +
  
  geom_line(aes(y = pred_poly, color = 'Polynomial Model'), linewidth = 1.5, alpha=0.35) +
  geom_point(aes(y = pred_poly, color = 'Polynomial Model'), size = 2, alpha=0.5) +
  
  scale_x_continuous(breaks = df_hiv_pred$YEAR) +
  scale_color_manual(values = c('Actual' = 'black',
                                'Linear Model' = 'blue',
                                'Polynomial Model' = 'red')) +
  labs(
    title = 'Actual vs Predicted HIV Diagnoses (Linear vs Polynomial)',
    x = 'Year',
    y = 'HIV Diagnoses',
    color = ''
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.95))
```

This plot compares actual HIV diagnoses with the linear and polynomial model predictions extended
through 2023. While both models stay close to the observed data from 2016 to 2021, their
extrapolated predictions diverge afterward. The linear model continues a sharp downward decline,
whereas the polynomial model flattens out and predicts a slower decrease. The actual values for 2022
and 2023 fall above both prediction lines, which is somewhat expected given the small sample size
and the disruptions caused by COVID-19 during the recent years.

### Geographic Exploration

We also explored the trends geographically, examining how HIV diagnoses varied across New York
City’s boroughs to understand whether the overall pattern differed by location.

#### Borough-level Model

We also fit a regression model using borough and year to examine how HIV diagnosis rates differ
across locations and how those borough-level trends have changed over time.

```{r}
borough_df_agg <- df %>% 
  filter(
    Borough != "All",
    Neighborhood == "All",
   SEX == "All",
   `RACE/ETHNICITY` == "All"
  )

# make a model
borough_agg_model = lm(formula =`HIV DIAGNOSES PER 100,000 POPULATION` ~ YEAR + Borough,
                           data = borough_df_agg)

# add predicted values to df
borough_df_agg$gen_pred_hiv_100 <- predict(borough_agg_model,
                                          newdata = borough_df_agg)

summary(borough_agg_model)
```

The model fits the borough-level data very well. The low residual standard error (2.755) indicates
strong accuracy, and the high R-squared (0.9301) and adjusted R-squared confirm that year and
borough together explain most of the variation in HIV diagnosis rates. The small p-value for the overall
F-statistic shows that these predictors jointly contribute significantly to the model. Each borough also
has a very low p-value, meaning their average rates differ significantly from one another. Combined
with the negative year coefficient, the results show two clear patterns: HIV diagnoses are declining
over time, and each borough follows its own distinct rate level.

```{r}
ggplot(data = borough_df_agg,
       aes(YEAR,`HIV DIAGNOSES PER 100,000 POPULATION`,
           color = Borough)) +
  geom_point() + 
  geom_line(aes(y = gen_pred_hiv_100)) +
  labs(
    title = "HIV Diagnoses by Borough",
    x = "",
    y = "HIV Diagnoses (per 100,000)"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

```{r}
borough_df_agg$residuals <- residuals(borough_agg_model)

ggplot(borough_df_agg, aes(x = gen_pred_hiv_100, y = residuals,
                           color = Borough)) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Predicted Values (Borough Model)",
    x = "Predicted HIV Diagnoses (per 100,000)",
    y = "Residuals"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

The residual plot shows no strong pattern, indicating that the borough model generally meets linearity
assumptions. Most residuals are close to zero, though Staten Island and the Bronx show larger
positive or negative values, indicating the model tends to underpredict or overpredict in those
boroughs. This suggests that while the model captures the overall trend, it struggles more with
boroughs that have unusually low or high HIV rates.

#### Neighborhood-level Model

We also fit a neighborhood-level model within the Bronx to examine how HIV diagnosis rates vary
across different areas of the borough and how those rates have changed over time.

```{r}
bronx_neighborhood_simple <- df %>% 
  filter(
    Borough == "Bronx",
    Neighborhood != "All",
   SEX == "All",
   `RACE/ETHNICITY` == "All")

bronx_neighborhood_model = lm(formula =`HIV DIAGNOSES PER 100,000 POPULATION` ~YEAR + Neighborhood, 
                   data = bronx_neighborhood_simple)

# add predicted values to df
bronx_neighborhood_simple$predicted_hiv_100 <- predict(bronx_neighborhood_model,
                                          newdata = bronx_neighborhood_simple)

summary(bronx_neighborhood_model)
```

The model indicates a clear downward trend in HIV diagnosis rates within the Bronx, with the year
coefficient showing a significant decline over time (p < 0.001). The overall model also fits well,
reflected by a high R-squared value (0.8212), meaning year and neighborhood together explain a
substantial share of the variation. The small p-value for the F-statistic confirms that the predictors
collectively contribute significantly to the model. Several neighborhoods have very low p-values,
indicating their average rates are statistically different from the reference neighborhood.
Overall, Bronx neighborhoods do not follow a single uniform pattern. Some areas have consistently
lower rates than others, even after accounting for the overall citywide decline.

```{r}
ggplot(data = bronx_neighborhood_simple,
       aes(YEAR,`HIV DIAGNOSES PER 100,000 POPULATION`,
           color = Neighborhood)) +
  geom_point(alpha = 0.8) +
  geom_line(aes(y = predicted_hiv_100)) +
  labs(
    title = "Yearly HIV Diagnoses by Bronx Neighborhood",
    x = "",
    y = "HIV Diagnoses (per 100,000)"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

```{r}
bronx_neighborhood_simple$residuals <- residuals(bronx_neighborhood_model)

ggplot(bronx_neighborhood_simple, aes(x = predicted_hiv_100, y = residuals,
                                      color = Neighborhood)) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Predicted HIV Diagnoses (per 100,000)",
    y = "Residuals"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```

The neighborhood-level residual plot shows a wider spread than the borough model, which means the
linear model has more difficulty capturing variation across individual neighborhoods. While most
points hover around zero, several neighborhoods show large positive or negative residuals, indicating
noticeable underprediction or overprediction. This suggests that the relationship between predictors
and HIV rates is more complex at the neighborhood level than a simple linear model can capture.

### Why We Moved Beyond Linear Models

After running these models, we realized that this dataset is not a great fit for standard linear modeling.
The dataset contains too few time points for a linear trend to be dependable. Borough and
neighborhood data also follow patterns driven more by population differences and reporting rules than
by smooth numerical trends. Because of this, the linear models did not capture the data’s structure
very well. So, we decided to treat the data differently and try a different type of regression that better
fits how this dataset behaves.

### A Different Approach - Tree-based Regression

Because linear models were not giving us meaningful results, we moved to tree-based methods,
starting with decision tree regression and then random forest regression. These models work
differently: instead of trying to fit one line to all the data, they split the data into smaller, more similar
groups and make predictions within those groups. This makes them more flexible and better suited for
patterns that are not linear. Random forests in particular average many trees together, which usually
gives more stable and accurate predictions. These methods allowed us to explore borough, race, and
demographic factors in a way that better matched the structure of the dataset.

To prepare the data for tree-based regression, we first created a modeling dataset containing the HIV
diagnosis rate along with key predictors such as year, borough, neighborhood, sex, race/ethnicity, and
the total number of diagnoses. All categorical variables were converted into factors so that the tree
model could correctly treat them as groups rather than numeric values. We then split the data into
training and testing sets using an 80–20 split to allow for model evaluation on unseen data. After
setting a random seed for reproducibility, we fit a decision tree regression model using the rpart
package.

### Decision Tree Regressor

We fit a decision tree regression model using the rpart function, with the HIV diagnosis rate as the
outcome and year, borough, neighborhood, sex, and race/ethnicity as predictors. The model uses the
"anova" method, which is designed for continuous outcomes. This approach lets the tree
automatically find splits in the data that best explain differences in HIV rates across groups and
locations. By doing so, it identifies which factors and combinations of factors contribute most to
variations in diagnosis rates.

```{r}
target = 'HIV DIAGNOSES PER 100,000 POPULATION'

model_df = df_no_agg[, c('HIV DIAGNOSES PER 100,000 POPULATION',
                  'YEAR',
                  'Borough',
                  'Neighborhood',
                  'SEX',
                  'RACE/ETHNICITY',
                  'TOTAL NUMBER OF HIV DIAGNOSES')]

model_df$Borough = factor(model_df$Borough)
model_df$Neighborhood = factor(model_df$Neighborhood)
model_df$SEX = factor(model_df$SEX)
model_df$`RACE/ETHNICITY` = factor(model_df$`RACE/ETHNICITY`)
```

```{r}
set.seed(65)
index = sample(1:nrow(model_df), 0.8 * nrow(model_df))

train = model_df[index, ]
test  = model_df[-index, ]
```

```{r}
set.seed(65)
tree_model = rpart(
  formula = `HIV DIAGNOSES PER 100,000 POPULATION` ~ YEAR + Borough + Neighborhood + SEX + `RACE/ETHNICITY`,
  data = train,
  method = 'anova'
)
```

```{r}
compact_split <- function(x, labs, digits, varlen, faclen) {
  sapply(labs, function(l) {
    
    feature <- sub("(.*?)(<|>|=).*", "\\1\\2", l)   # keeps "Neighborhood =" or "TOTAL <"
    feature <- trimws(feature)
    
    values <- sub(".*?(<|>|=)", "", l)  # remove feature+operator
    values <- trimws(values)
    
    parts <- unlist(strsplit(values, ",")) 
    parts <- trimws(parts)
    
    keep_n <- 3
    
    keep <- parts[1:min(keep_n, length(parts))]
    leftover <- length(parts) - keep_n
    
    if (leftover > 0) {
      keep <- c(keep, paste0("(", leftover, " more)"))
    }
    
    paste(
      feature,
      paste(keep, collapse = "\n"),
      sep = "\n"
    )
  })
}
```

```{r fig.width=16}
rpart.plot(
  tree_model,
  type = 1,
  extra = 1,
  faclen = 0,
  cex = 0.6,
  split.fun = compact_split
)
```

The decision tree begins with race and ethnicity because this variable creates the strongest
separation in the data. As the tree moves downward, neighborhood and sex appear often because
they help refine the groups even further. Each final leaf represents a subgroup with its own average
predicted HIV rate, showing how combinations of demographic and geographic factors shape the
overall pattern. The structure of the tree makes it clear that the differences in HIV rates are not driven
by any single factor but by how these characteristics interact with one another.

```{r}
imp <- as.data.frame(tree_model$variable.importance)
imp$Feature <- rownames(imp)
colnames(imp)[1] <- 'Importance'

ggplot(imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = 'steelblue') +
  coord_flip() +
  labs(
    title = 'Feature Importance - Decision Tree',
    x = 'Feature',
    y = 'Importance Score'
  ) +
  theme_minimal() +
  theme(
      plot.title = element_text(size = 14, face = 'bold'),
      axis.text  = element_text(size = 11)
    )
```

The feature importance results show that Neighborhood contributes the most to predicting HIV
diagnosis rates in the decision tree. Race/ethnicity is the next most important factor, followed by sex.
Borough has the smallest impact, likely because neighborhood information already captures most of
the geographic variation. Overall, this tells us that local neighborhood differences explain more of the
variation in HIV rates than broader borough-level or demographic factors alone.

```{r fig.width=8, fig.height=4}
plotcp(tree_model)
```

This plot shows how the tree’s cross-validated error changes as the model becomes more complex.
When the tree is very small, the error is high because the model cannot capture much of the variation.
As the tree grows, the error drops steadily, meaning additional splits help improve prediction
accuracy. After a certain point, the curve levels off, which tells us that making the tree larger provides
only small improvements. This pattern helps identify a reasonable tree size that improves accuracy
without becoming unnecessarily complex.

```{r}
best_cp <- 0.017

#best_cp <- hiv_tree$cptable[
#  which.min(hiv_tree$cptable[, "xerror"]),
#  "CP"
#]

pruned_tree = prune(tree_model, cp = best_cp)
```


```{r}
eval_metrics <- function(m, test) {
  pred   <- predict(m, test)
  actual <- test$`HIV DIAGNOSES PER 100,000 POPULATION`
  MAE  <- round(mean(abs(pred - actual)), 2)
  MSE  <- round(mean((pred - actual)^2), 2)
  RMSE <- round(sqrt(MSE), 2)
  R2   <- round(1 - sum((pred - actual)^2) / sum((actual - mean(actual))^2), 2)
  c(MAE, MSE, RMSE, R2)
}

metrics_df <- data.frame(
  Metric      = c('MAE','MSE','RMSE','R²'),
  Decision_Tree = eval_metrics(tree_model, test),
  Pruned_Tree = eval_metrics(pruned_tree, test)
)

flextable(metrics_df) |>
  set_caption("Decision Tree Vs Pruned Tree (Cp = 0.017)")
```

The full decision tree and the pruned tree perform very similarly. The pruned tree shows a small increase 
in MAE but achieves slightly lower MSE and RMSE, indicating marginally better average error overall. 
Its R² increases from 0.46 to 0.48, meaning the pruned model explains a little more variance in HIV 
diagnosis rates. In this case, pruning produces a simpler model without hurting performance and may even 
offer a modest improvement.

### Random Forest Regressor

We trained a random forest model to improve prediction accuracy. Unlike a single decision tree, a
random forest builds many trees using random subsets of the data and then averages their
predictions, which helps reduce overfitting and produce more stable results. We used 500 trees and
set mtry to 3, allowing the model to consider three predictors at each split. This approach lets us
capture more complex patterns in HIV rates across neighborhoods and demographic groups.

```{r}
model_df <- subset(model_df, select = -c(`TOTAL NUMBER OF HIV DIAGNOSES`))
```

```{r}
set.seed(65)
train_idx <- sample(seq_len(nrow(model_df)), size = 0.8 * nrow(model_df))

train <- model_df[train_idx, ]
test  <- model_df[-train_idx, ]

y_train <- train$`HIV DIAGNOSES PER 100,000 POPULATION`
x_train <- train[, setdiff(names(train), "HIV DIAGNOSES PER 100,000 POPULATION")]

y_test  <- test$`HIV DIAGNOSES PER 100,000 POPULATION`
x_test  <- test[, setdiff(names(test), "HIV DIAGNOSES PER 100,000 POPULATION")]

rf_model <- randomForest(
  x = x_train,
  y = y_train,
  ntree = 500,
  mtry = 3,
  importance = TRUE
)

rf_vals <- eval_metrics(rf_model, test)

metrics_df$Random_Forest <- rf_vals

flextable(metrics_df) |>
  set_caption("Decision Tree Vs Pruned Tree Vs Random Forest")
```

The random forest model performs noticeably better than both the full and pruned decision trees. It
has the lowest MAE, MSE, and RMSE values, meaning its predictions are closer to the actual HIV
rates. Its R² value is also much higher at 0.65, indicating that it explains a larger share of the variation
in the data. These results show that combining many trees produces a more accurate and stable
model than relying on a single tree, which is expected given the complexity and variability of HIV rates
across neighborhoods and demographic groups.

```{r fig.width=8, fig.height=3}
imp <- importance(rf_model)

imp_df <- data.frame(
  Feature = rownames(imp),
  Importance = imp[ , 1]
)
rownames(imp_df) <- NULL

ggplot(imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = '#2c7fb8') +
  coord_flip() +
  labs(
    title = 'Feature Importance - Random Forest (%IncMSE)',
    x = 'Feature',
    y = 'Importance'
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, face = 'bold'),
    axis.text  = element_text(size = 10)
  )
```

The random forest feature importance results show that race and ethnicity is the strongest predictor of
HIV diagnosis rates, followed closely by sex. Neighborhood still plays an important role but
contributes less than the first two variables. Borough and year have much smaller impacts once the
other predictors are included. This pattern highlights demographic characteristics as the main drivers
of HIV rate variation in the random forest model. In the decision tree, neighborhood appeared as the
most important factor, which reflects how a single tree often relies on large geographic splits. The
random forest spreads its importance across many trees, which brings demographic variables like
race and sex to the forefront.

### RF Cross-Validation

To tune the random forest model, we used 10-fold cross-validation and tested different combinations
of the number of trees (ntree) and the number of predictors considered at each split (mtry). For each
ntree value, the model evaluated five possible mtry settings and selected the one that produced the
lowest RMSE. This process allows the model to choose the tuning parameters that give the best
predictive performance rather than relying on defaults. After collecting the results across all ntree
values, we compared them and identified the combinations with the highest R² and lowest errors. This
cross-validation step helps ensure that the final random forest model is well-optimized and
generalizes better to unseen data.

```{r}
control <- trainControl(
  method = 'cv',
  number = 10
)

set.seed(65)

results_list <- list()

for (m in c(200, 500, 800, 1000, 1500, 2000)){

  rf_cv <- train(
    `HIV DIAGNOSES PER 100,000 POPULATION` ~ .,
    data = train,
    method = 'rf',
    trControl = control,
    tuneGrid = data.frame(mtry = c(1,2,3,4)),
    ntree = m
  )

  best <- rf_cv$results[which.min(rf_cv$results$RMSE), ]

  results_list[[length(results_list) + 1]] <- data.frame(
    ntree    = m,
    mtry     = best$mtry,
    RMSE     = round(best$RMSE, 2),
    Rsquared = round(best$Rsquared, 2),
    MAE      = round(best$MAE, 2)
  )
}

results_df <- do.call(rbind, results_list)
results_sorted_desc <- results_df[order(-results_df$Rsquared), ]
flextable(head(results_sorted_desc, 5)) |>
  set_caption("Top 5 Cross-Validated Scores - Random Forest")
```

The cross-validation results show that the random forest performs consistently well across different numbers of trees, with RMSE values around 29 and R² values in the range of 0.59 to 0.61. This suggests that, on average, the model can explain a little more than half of the variation in HIV diagnosis rates when evaluated on different subsets of the training data. 

The initial random forest model trained with 500 trees and mtry = 3 achieved a higher R² of about 0.65 on the test set, which reflects the performance on that specific data split. This difference can occur when the test data happen to follow patterns that the model learned especially well, making that particular split slightly easier to predict than the folds used during cross-validation.

### Can a Random Forest Trained on Past Data Predict the Future?

To explore this question, we trained the model using data from 2016-2020 and evaluated its
performance on the unseen year 2021. Although Random Forests are not traditional forecasting
models and YEAR itself was not a strong predictor, this temporal split allows us to assess whether the
demographic and geographic patterns learned in earlier years remain stable when applied to a new
period.

The goal is not to forecast future values in the strict time-series sense, but to test the model’s ability to generalize across years.

```{r}
df2 <- model_df
```

```{r}
# cutoff year
cutoff_year <- 2021

# split data based on the year column
time_train <- df2[df2$YEAR < cutoff_year, ]
time_test <- df2[df2$YEAR == cutoff_year, ]
```

```{r}
# separate features and target
y_time_train <- time_train$`HIV DIAGNOSES PER 100,000 POPULATION`
x_time_train <- time_train[, setdiff(names(time_train), "HIV DIAGNOSES PER 100,000 POPULATION")]

y_time_test <- time_test$`HIV DIAGNOSES PER 100,000 POPULATION`
x_time_test <- time_test[, setdiff(names(time_test), "HIV DIAGNOSES PER 100,000 POPULATION")]
```

```{r}
set.seed(65)

# retrain random forest model
rf_time_model <- randomForest(
  x = x_time_train,
  y = y_time_train,
  ntree = 500,
  mtry = 2,
  importance = TRUE
)

# generate predictions for 2021
rf_time_pred <- predict(rf_time_model, newdata = x_time_test)
```

```{r}
# calculating errors
residuals_time <- rf_time_pred - y_time_test
MAE_time <- mean(abs(residuals_time))
MSE_time <- mean(residuals_time^2)
RMSE_time <- sqrt(MSE_time)
SST_time <- sum((y_time_test - mean(y_time_test))^2)
SSE_time <- sum(residuals_time^2)                   
R2_time <- 1 - (SSE_time / SST_time)

metrics_time_df <- data.frame(
  mtry = 2,
  ntree = 500,
  MAE  = MAE_time,
  MSE  = MSE_time,
  RMSE = RMSE_time,
  R2   = R2_time
)

metrics_time_df <- round(metrics_time_df, 3)

flextable(metrics_time_df) |>
  set_caption("RF Model Evaluation on Unseen Year (2021)")
```

The model performs well on the held-out data. With its chosen parameters, it reaches an MAE of about 11 and an RMSE of about 24, meaning the forecasts for 2021 are generally close to the true HIV rates. The R² value of 0.612 shows that the model captures over 60 percent of the variation, which indicates strong year-to-year generalization.

### Random Forest Hyperparameter Search

We ran a grid-search experiment by tuning four random forest parameters: mtry (1 to 4), ntree (200 to 1000), nodesize (3 to 15), and maxnodes (15 to 50). For each combination we trained a model and recorded the out-of-bag RMSE and R² to assess performance. Sorting the results by R² helped identify the strongest parameter settings.

```{r, include=FALSE}
mtry_values    <- c(1, 2, 3, 4)
ntree_values   <- c(200, 500, 800, 1000)
nodesize_values <- c(3, 5, 10, 15)
maxnodes_values <- c(15, 25, 35, 50)

results3 <- data.frame(
  mtry = integer(),
  ntree = integer(),
  nodesize = integer(),
  maxnodes = integer(),
  RMSE = numeric(),
  R2 = numeric()
)

set.seed(65)

for (m in mtry_values) {
  for (n in ntree_values) {
    for (ns in nodesize_values) {
      for (mn in maxnodes_values) {

        model <- randomForest(
          x = x_time_train,
          y = y_time_train,
          mtry = m,
          ntree = n,
          nodesize = ns,
          maxnodes = mn,
          importance = FALSE
        )

        # Out-of-bag predictions
        oob_pred <- model$predicted
        actual   <- y_time_train

        valid <- !is.na(oob_pred)
        oob_pred <- oob_pred[valid]
        actual   <- actual[valid]

        # Compute RMSE and R²
        rmse <- sqrt(mean((oob_pred - actual)^2))
        r2   <- 1 - sum((oob_pred - actual)^2) / sum((actual - mean(actual))^2)

        # Save results
        results3 <- rbind(
          results3,
          data.frame(
            mtry = m,
            ntree = n,
            nodesize = ns,
            maxnodes = mn,
            RMSE = rmse,
            R2 = r2
          )
        )
      }
    }
  }
}

results3[order(-results3$R2), ]
```

```{r}
# sort the results by OOB R-squared in descending order and select the top 3 rows
top_3_params <- results3[order(-results3$R2), ][1:3, ]

flextable(top_3_params)
```

We then trained a final random forest model using the best parameter combination identified in the grid search. The selected settings were mtry = 3, ntree = 800, nodesize = 10, and maxnodes = 50. After fitting this model on the training years, we used it to generate predictions for the 2021 test data to assess how well the tuned model performs on unseen observations.

```{r}
# parameters for the best model
best_params <- top_3_params[1, ]

set.seed(65)

# train model
rf_best_model <- randomForest(
  x = x_time_train,
  y = y_time_train,
  mtry = best_params$mtry,
  ntree = best_params$ntree,
  nodesize = best_params$nodesize,
  maxnodes = best_params$maxnodes,
  importance = FALSE 
)

# predict the 2021 test data
pred_best <- predict(rf_best_model, newdata = x_time_test)
```

```{r}
residual_data <- data.frame(
  Actual = y_time_test,
  Predicted = pred_best,
  Residuals = y_time_test - pred_best # Actual - Predicted
)

# residuals vs predicted values
ggplot(residual_data, aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.6, color = 'darkblue') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'red', linewidth = 1) +
  labs(
    title = paste("Residual Plot for Random Forest (Best RF Model)"),
    subtitle = "HIV Diagnoses Per 100,000 Population (Year 2021)",
    x = "Predicted HIV Rate",
    y = "Residuals (Actual - Predicted)"
  ) +
  theme_minimal()
```

The residual plot shows that the model predicts lower HIV diagnosis rates fairly accurately, with most points near zero for predicted values under 50. This happens because most observations fall in this range, giving the model more training examples. At higher predicted rates, the residuals spread out and become more extreme, meaning the model is less reliable for neighborhoods or groups with very high HIV diagnosis rates, which are much less common in the data.


```{r}
ggplot(residual_data, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, color = 'forestgreen') +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'red', linewidth = 1) +
  labs(
    title = "Actual vs. Predicted Rates",
    subtitle = "Forecasted 2021 HIV Rates (Per 100,000 Population)",
    x = "Actual HIV Rate (2021)",
    y = "Predicted HIV Rate (2021)"
  ) +
  theme_minimal()
```

The plot shows that as the actual HIV rate increases, the model’s predicted rate also increases, so it is capturing the overall relationship in the data. For most observations under about 50 cases per 100,000, the predictions stay close to the diagonal reference line. At higher actual rates, many points fall below the line, indicating that the model underpredicts these extreme cases. This aligns with what we observed in the residual plot, where larger positive residuals appeared mostly at higher predicted values, confirming that the model struggles with the rare high-rate observations and tends to shrink them toward the center of the distribution.

```{r}
set.seed(65)
k <- 10
folds <- sample(rep(1:k, length.out = nrow(x_time_train)))

cv_MAE  <- c()
cv_MSE  <- c()
cv_RMSE <- c()
cv_R2   <- c()

for(i in 1:k){
  
  train_idx <- folds != i
  test_idx  <- folds == i
  
  rf_cv <- randomForest(
    x = x_time_train[train_idx, ],
    y = y_time_train[train_idx],
    mtry     = best_params$mtry,
    ntree    = best_params$ntree,
    nodesize = best_params$nodesize,
    maxnodes = best_params$maxnodes
  )
  
  pred <- predict(rf_cv, x_time_train[test_idx, ])
  actual <- y_time_train[test_idx]
  
  residuals <- pred - actual
  
  cv_MAE[i]  <- mean(abs(residuals))
  cv_MSE[i]  <- mean(residuals^2)
  cv_RMSE[i] <- sqrt(cv_MSE[i])
  cv_R2[i]   <- 1 - sum(residuals^2) / sum((actual - mean(actual))^2)
}

cv_results <- data.frame(
  mtry     = best_params$mtry,
  ntree    = best_params$ntree,
  nodesize = best_params$nodesize,
  maxnodes = best_params$maxnodes,
  MAE      = mean(cv_MAE),
  MSE      = mean(cv_MSE),
  RMSE     = mean(cv_RMSE),
  R2       = mean(cv_R2)
)

flextable(round(cv_results, 3)) |>
  set_caption("Cross-Validation - RF Model with Best Parameters")
```

The cross-validation result for the model with mtry = 3, ntree = 800, nodesize = 10, and maxnodes = 50 gives an R² of about 0.615. This is very close to the training R² of about 0.612 from the initial model with mtry = 2 and ntree = 500, which we used to predict on unseen 2021 data. The similarity between these values suggests that the model stays stable during tuning and does not show clear signs of overfitting.

### AIDS Rate Modeling

We also extended our analysis by trying to predict AIDS diagnoses per 100,000 population using both
decision trees and random forests. The models followed the same setup as our HIV rate experiments.

```{r}
hiv <- df_no_agg

hiv <- hiv %>%
  mutate(
    year = factor(YEAR),
    borough = factor(Borough),
    neighborhood = factor(Neighborhood),
    sex = factor(SEX),
    race_ethnicity = factor(`RACE/ETHNICITY`)
  ) %>%
  rename(
    aids_diagnoses_per_100_000_population =
      `AIDS DIAGNOSES PER 100,000 POPULATION`
  )

rmse <- function(obs, pred) {
  sqrt(mean((obs - pred)^2))
}
```

```{r}
set.seed(65)

train_index <- createDataPartition(
  hiv$aids_diagnoses_per_100_000_population,
  p = 0.8,
  list = FALSE
)

hiv_train <- hiv[train_index, ]
hiv_test  <- hiv[-train_index, ]

aids_formula <- aids_diagnoses_per_100_000_population ~
  year + borough + neighborhood + sex + race_ethnicity
```

```{r}
hiv_tree <- rpart(
  formula = aids_formula,
  data = hiv_train,
  method = "anova",
  control = rpart.control(cp = 0.01)
)

imp <- as.data.frame(hiv_tree$variable.importance)
imp$Feature <- rownames(imp)
colnames(imp)[1] <- 'Importance'

ggplot(imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = 'steelblue') +
  coord_flip() +
  labs(
    title = 'Feature Importance - Decision Tree',
    x = 'Feature',
    y = 'Importance Score'
  ) +
  theme_minimal() +
  theme(
      plot.title = element_text(size = 14, face = 'bold'),
      axis.text  = element_text(size = 11)
    )
```

The decision tree importance plot shows that neighborhood is the strongest predictor of AIDS
diagnoses, followed by race/ethnicity and sex. Borough contributes less, and year has very little
influence. This means the model relies most on local neighborhood-level differences when making
predictions.

```{r}
plotcp(hiv_tree)
```

The cross-validation curve shows a steady reduction in relative error as cp decreases, and the
minimum point occurs around cp ≈ 0.012. This is where the tree achieves its lowest cross-validated
error, indicating the best balance between model complexity and predictive performance for this
dataset.

```{r}
best_cp <- 0.023

hiv_tree_pruned <- prune(hiv_tree, cp = best_cp)

eval_metrics <- function(m, test) {
  pred   <- predict(m, test)
  actual <- test$aids_diagnoses_per_100_000_population
  MAE  <- round(mean(abs(pred - actual)), 2)
  MSE  <- round(mean((pred - actual)^2), 2)
  RMSE <- round(sqrt(MSE), 2)
  R2   <- round(1 - sum((pred - actual)^2) / sum((actual - mean(actual))^2), 2)
  c(MAE, MSE, RMSE, R2)
}

metrics_df <- data.frame(
  Metric      = c('MAE','MSE','RMSE','R²'),
  Decision_Tree = eval_metrics(hiv_tree, hiv_test),
  Pruned_Tree = eval_metrics(hiv_tree_pruned, hiv_test)
)

flextable(metrics_df) |>
  set_caption("Decision Tree Vs Pruned Tree (Cp = 0.025)")
```

Even though the cross-validation curve identifies cp ≈ 0.012 as the best value, the tree pruned at cp =
0.025 achieved slightly better RMSE, and R² on our evaluation set. This happens because rpart
optimizes cross-validated relative error, which does not always align perfectly with external
performance metrics, especially in smaller or noisier datasets. A slightly simpler tree can generalize
better in practice.

```{r plot-tree, fig.width=10, fig.height=6}
par(xpd = NA, mar = c(1, 1, 3, 1))

rpart.plot(
  hiv_tree_pruned,
  type = 2,
  extra = 101,
  fallen.leaves = TRUE,
  tweak = 1.2,
  cex = 0.6,
  faclen = 6,
  main = "Pruned Decision Tree",
  split.fun = compact_split
)
```

The pruned tree shows that neighborhood is the dominant factor separating HIV diagnosis rates, with
race and ethnicity and sex providing additional refinements within each group. The simplified structure
still captures the same hierarchy seen earlier, and it further reinforces the feature importance results,
where neighborhood and demographic variables were the strongest predictors.

```{r}
set.seed(65)

cv_ctrl <- trainControl(
  method = "cv",
  number = 10
)

tree_cv <- train(
  aids_formula,        # formula as first argument (no "formula =")
  data = hiv_train,
  method = "rpart",
  trControl = cv_ctrl,
  tuneLength = 5
)

tree_results <- tree_cv$results[ , c('cp', 'RMSE', 'MAE', 'Rsquared')]

tree_results <- round(tree_results, 3)

flextable(tree_results) |>
  set_caption('Cross-Validated Scores - Decision Tree')
```

The cross-validation results show that the decision tree performs best at cp = 0.020, where RMSE
and MAE are lowest and R² is highest at about 0.223. As cp increases, the model becomes more
aggressively pruned and performance worsens, with R² dropping and errors rising. This indicates that
lightly pruned trees capture the AIDS-rate patterns better than more simplified versions, though
overall predictive power remains modest.

### Random Forest

We also fit a random forest model to predict AIDS diagnoses per 100,000, using 500 trees and mtry =
3. This model provides more stable predictions than a single tree and again highlights neighborhood
and demographic variables as the strongest predictors.

```{r}
set.seed(65)
hiv_rf <- randomForest(
  formula = aids_formula,
  data = hiv_train,
  ntree = 500,
  mtry = 3,
  importance = TRUE,
  na.action = na.omit
)

importance_df <- importance(hiv_rf)
importance_df <- as.data.frame(importance_df)

importance_df$variable <- rownames(importance_df)

importance_df %>%
  arrange(desc(`%IncMSE`)) %>%
  ggplot(aes(x = reorder(variable, `%IncMSE`), y = `%IncMSE`)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Feature Importance - Random Forest (AIDS rate)",
    x = "Predictor",
    y = "% increase in MSE"
  ) +
  theme_minimal()
```

The random forest feature importance results show that race and ethnicity is the strongest predictor of
AIDS rates, followed closely by sex. Neighborhood also has a meaningful impact but contributes less
than the demographic variables. Borough and year add very little once the model accounts for race,
sex, and neighborhood.

```{r}
metrics_df <- data.frame(
  Metric      = c('MAE','MSE','RMSE','R²'),
  Decision_Tree = eval_metrics(hiv_tree, hiv_test),
  Pruned_Tree = eval_metrics(hiv_tree_pruned, hiv_test),
  Random_Forest = eval_metrics(hiv_rf, hiv_test)
)

flextable(metrics_df) |>
  set_caption("Decision Tree Vs Pruned Tree Vs Random Forest")
```

The results show that predicting AIDS diagnoses is more challenging than predicting HIV rates. All
three models have relatively low R² values, meaning they explain only a small portion of the variation
in AIDS rates. The random forest performs the best, with the lowest MAE and RMSE and the highest
R² at 0.30, but the improvement over the decision tree models is modest. Overall, the models capture
some structure in the data, but AIDS rates appear more difficult to model accurately.

```{r}
set.seed(65)
rf_cv <- train(
  aids_formula,
  data = hiv_train,
  method = "rf",
  trControl = cv_ctrl,
  tuneLength = 5,
  importance = TRUE
)

rf_results <- rf_cv$results[ , c('mtry', 'RMSE', 'MAE', 'Rsquared')]

rf_results <- round(rf_results, 3)

rf_results_sorted <- rf_results[order(-rf_results$Rsquared), ]

flextable(rf_results_sorted) |>
  set_caption('Cross-Validated Scores - Random Forest')
```

The cross-validation results for the random forest model show that performance improves noticeably
as mtry increases. The best scores appear around mtry = 15-28, where RMSE is roughly 26 and R²
reaches about 0.40. This means the model explains around 40 percent of the variation during cross-
validation, which is stronger than the decision tree results.

```{r}
tree_train_pred <- predict(hiv_tree_pruned, newdata = hiv_train)
tree_test_pred  <- predict(hiv_tree_pruned, newdata = hiv_test)

rf_train_pred <- predict(hiv_rf, newdata = hiv_train)
rf_test_pred  <- predict(hiv_rf, newdata = hiv_test)

performance_df <- tibble(
  model = c("Decision tree (pruned)", "Random forest"),
  rmse_train = c(
    rmse(hiv_train$aids_diagnoses_per_100_000_population, tree_train_pred),
    rmse(hiv_train$aids_diagnoses_per_100_000_population, rf_train_pred)
  ),
  rmse_test = c(
    rmse(hiv_test$aids_diagnoses_per_100_000_population, tree_test_pred),
    rmse(hiv_test$aids_diagnoses_per_100_000_population, rf_test_pred)
  )
)

rf_resid <- hiv_test$aids_diagnoses_per_100_000_population - rf_test_pred

resid_df <- tibble(
  fitted = rf_test_pred,
  residual = rf_resid
)

ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.6) +
  labs(
    title = "Residuals vs Fitted Values - Random Forest (AIDS Rate)",
    x = "Predicted AIDS diagnoses per 100,000",
    y = "Residual"
  ) +
  theme_minimal()
```

The residual plot shows that the random forest predicts lower AIDS rates fairly well, but the errors get
larger as the predicted values increase. Most points near 0-30 diagnoses per 100,000 cluster close to
the horizontal line, meaning good accuracy in the common range. At higher predicted values, the
residuals spread out more and often fall below zero, which indicates underprediction for the highest-
rate neighborhoods. This matches what we saw earlier with HIV rates: the model performs best where
the data is dense and struggles with rare, extreme cases.

```{r}
rf_results <- tibble(
  actual    = hiv_test$aids_diagnoses_per_100_000_population,
  predicted = rf_test_pred
)

ggplot(rf_results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Random Forest - Predicted vs Actual AIDS Diagnosis Rates",
    x = "Actual AIDS Rate per 100,000",
    y = "Predicted AIDS Rate per 100,000"
  ) +
  theme_minimal()
```

The predicted vs actual plot shows that the random forest model captures the general increase in
AIDS rates but consistently underestimates the highest values. Most low-rate observations (near zero)
are predicted reasonably well, but as the actual AIDS rate rises, the predicted values fall below the
diagonal reference line. This underprediction at the upper end matches what we saw in the residual
plot, where large positive residuals appeared for high-rate neighborhoods.

## Limitations

Our analysis has several limitations that should be considered when interpreting the results. First, the
dataset contains only six years of consistently reported values for many of the variables, which limits
the reliability of trend modeling and reduces the statistical power of time-based comparisons. Many
neighborhood and demographic categories also have sparse data, particularly for smaller populations,
which restricts model performance and contributes to the underprediction of extreme values. In
addition, the structure of the surveillance dataset reflects reporting practices rather than a randomly
sampled process, meaning some patterns arise from data availability rather than underlying
epidemiology. Finally, tree-based models capture group differences well but are less effective for
forecasting rare, high-burden observations, since the data are highly imbalanced and dominated by
low-rate neighborhoods.

## Future Works

• Incorporate additional years of data and bring in external variables such as socioeconomic
indicators, access-to-care metrics, or viral suppression rates to improve explanatory power.

• Explore more advanced models such as gradient boosting or spatial models that account for
geographic relationships between neighborhoods.

• Expand and refine the feature set to reduce missing information and improve model robustness.

• Build a more formal forecasting framework using rolling or expanding training windows to better
assess how models generalize to future years.

• Evaluate model stability across different boroughs and demographic groups to understand where
errors are highest and why.

## Wrap Up

Overall, the project provides a detailed view of how HIV and AIDS diagnoses vary across time,
boroughs, neighborhoods, and demographic groups in New York City. The EDA confirms persistent
geographic and demographic disparities, with the Bronx, Black and Latino/Hispanic populations, and
males showing consistently higher burdens.

Linear models captured broad trends but were not well suited for the structure of the dataset, leading
us to adopt tree-based methods. Decision trees offered interpretable groupings, while random forests
delivered stronger predictive performance and more stable generalization across years.

Even with their limitations, these models highlight the key factors shaping HIV and AIDS rates and
help identify which communities carry the greatest burden. Together, these findings reinforce the need
for targeted public health interventions and continued monitoring of neighborhood-level disparities.
